---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)
```

# cbr

<!-- badges: start -->
[![R-CMD-check](https://github.com/mdsumner/cbr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mdsumner/cbr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`cbr` is an experimental R package built on the
[controlledburn](https://crates.io/crates/controlledburn) Rust crate. It
rasterizes polygons, lines and points onto a regular grid and keeps the result
*sparse*: polygon interiors as horizontal runs of fully covered cells, boundary
cells with their exact coverage fraction, and nothing at all for the empty
cells. A dense raster is something you ask for at the end, if you need it.

## Status

Experimental. `cbr` explores the ideas in
[hypertidy/controlledburn](https://github.com/hypertidy/controlledburn) and
related hypertidy packages, but on a different foundation: the Rust crate is a
pure port of the C++ core, and `cbr` is a thin binding to it via
[extendr](https://extendr.github.io/). There is no connection to the
controlledburn R package's code or API, and no obligation to be compatible with
it. This is deliberately a place to try things; ideas that work here may be
folded back into the parent projects, where backwards compatibility matters
more.

Two conventions differ from the parent that are worth knowing up front:

* Grid extents are `c(xmin, xmax, ymin, ymax)`, as in `terra::ext()`.
* Materialized rasters are in the layout of `gdalraster::read_ds()`: a plain
  vector in row-major order from the top-left cell, with a `gis` attribute
  (`type`, `bbox`, `dim`, `srs`, `datatype`). It is a stable format that needs
  no dependency to produce or consume, and it plugs into `gdalraster` tooling
  directly.

## Installation

You need a Rust toolchain (`cargo` and `rustc`). Then

```r
# install.packages("pak")
pak::pak("mdsumner/cbr")
```

## The burn object

`cb_burn()` takes anything `wk::as_wkb()` understands and returns a `cb_burn`:
five data frames and the grid definition in attributes. Below `geom` is a set
of polygons in longitude-latitude.

```{r burn}
library(cbr)
## get some polygons in wk form however you like
geom <- geos::as_geos_geometry(geodata::world(path = tempdir()))
b <- cb_burn(geom)
b
```

Called with just the geometry, the grid is the bounding box of the input at 256
columns, with the number of rows chosen for square cells. Supply `extent`
and/or `dimension` to control it; a single `dimension` value is taken as the
number of columns.

```{r burn-grid}
b_fine <- cb_burn(geom, dimension = 2048)
b_fine
```

The tables are the whole result. `runs` holds the interiors (one record per
row per polygon per contiguous span), `edges` the boundary cells with their
coverage fraction, and `id` is the 1-based position of the geometry in the
input.

```{r tables}
head(b$runs)
head(b$edges)
```

## Summarize and plot without materializing

Everything in `summary()` costs time proportional to the number of records in
the tables, never to the number of cells: exact covered-cell totals per row and
per column, the extent of touched cells, the area of each geometry in cells,
and a coarse overview raster of mean coverage per tile.

```{r summary}
s <- summary(b_fine)
s
head(s$area)
```

`plot()` uses the same machinery. The main panel is either every record drawn
as a rectangle (`what = "exact"`, the picture of the sparse representation
itself) or the tiled overview (`what = "overview"`), chosen automatically by
record count. The strips along the top and right are the exact column and row
totals, so they stay exact even when the panel is an overview.

```{r plot, fig.width = 8, fig.height = 6}
plot(b_fine, main = "coverage")
```

```{r plot-id, fig.width = 8, fig.height = 6}
plot(b, what = "exact", by_id = TRUE, marginals = FALSE, main = "by geometry")
```

## Crop, then materialize

`cb_crop()` snaps an extent to the grid and subsets the tables to it. Nothing
is re-rasterized: runs that straddle the window are clipped and indices are
re-based, so the result is a valid burn on the smaller grid that remembers its
parent. With no `extent` it crops to the content.

```{r crop}
cr <- cb_crop(b_fine)
cr

cr1 <- cb_crop(b_fine, extent = c(100, 150, -45, -20))
cr1

cb_crop(cr1) ## we lost the sea of emptiness
plot(cr1)
```

`cb_materialize()` expands a burn to a dense vector in `read_ds` layout.
`value` chooses what a cell holds: summed `"coverage"` (Float32), the `"id"` of
the covering geometry (Int32, highest wins where geometries overlap), or the
`"count"` of geometries touching the cell.

```{r materialize}
v <- cb_materialize(cr)
str(v)

m <- cb_as_matrix(v)
dim(m)
```

Because the result is `read_ds`-shaped, `gdalraster` can plot it or write it
out with no conversion:

```{r gdalraster, eval = requireNamespace("gdalraster", quietly = TRUE), fig.width = 8, fig.height = 5}
gdalraster::plot_raster(v, legend = TRUE, main = "materialized coverage", col_map_fn = grey.colors(25))
```

`cb_overview()` is the same shape at a coarser resolution: mean coverage per
block of cells, computed from the tables with a difference array so that runs
are never expanded whatever the block size. With `block = 1` it equals
`cb_materialize()`.

```{r overview}
ov <- cb_overview(b_fine, block = 16)
str(ov)
```

## Why sparse

For a polygon layer the number of records is roughly `nrow * ngeom` for runs
plus the boundary length in cells for edges. The number of cells is
`ncol * nrow`. On a grid of a few thousand columns those differ by orders of
magnitude, and the gap widens as the grid gets finer:

```{r why-sparse}
s_fine <- summary(b_fine)
c(
  cells = prod(s_fine$dimension),
  records = sum(s_fine$records[c("runs", "edges")]),
  interior_cells = sum(s_fine$area$interior)
)
```

The tables answer most questions directly. Area per geometry, a row or column
profile, an overview at any resolution, and a crop to any window are all
rowsum-style operations on integer keys. Materializing is the one step that
touches every cell, and cropping first means it only touches the cells you
want.

## Workflows

* **Zonal mask**: `cb_burn(polys, extent, dimension)` then
  `cb_materialize(., "id")` gives the zone raster, or use the tables directly
  to index into a matching data raster by `(row, col)`.
* **Exact fractional coverage**: the default `coverage = TRUE`; sum
  `s$area$interior + s$area$boundary` for area in cells per geometry.
* **Fast mask**: `coverage = FALSE` for the cell-centre rule, runs only.
* **Big grid, small feature**: burn at full resolution, `cb_crop()` to the
  content, then materialize only that window.
* **Look before you allocate**: `summary()` and `plot()` on any burn, at any
  size.

## Related

* [controlledburn](https://crates.io/crates/controlledburn), the Rust crate.
* [hypertidy/controlledburn](https://github.com/hypertidy/controlledburn),
  the original R package with the C++ core.
* [wk](https://paleolimbot.github.io/wk/) for geometry input, and
  [gdalraster](https://usdaforestservice.github.io/gdalraster/) for the
  `read_ds` raster layout.

## Code of Conduct

Please note that the cbr project is released with a
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
